Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2021-11-04 Manuscript/Manuscript.Rmd
Add some more references to manuscript

- What is heterogeniety? I.e. variability within studies vs. variability between studies

@@

- What methods are required to analyze multilevel data?	Add references, e.g. @hox17. At least explain difference random effects for intercept term, predictor effects, and/or variance residual error.

@@

-	What kinds of missingness are there? ADD: missingness mechanisms here.

@@

- What types of multilevel methods are available? General overview of approaches, see @audi18 @grun18. E.g., imputation of study level versus patient-level covariates, and one-stage imputation versus two-stage imputation methods.

@@

- We focus on the workflow for conditional modeling (not JOMO) in `mice`. Refer to other packages: `mitml`, `miceadds`.

- Case study options: `metamisc::impact` (real data on traumatic brain injuries, IPD), `mice::popularity` (simulated data with MNAR/MAR mixture, schools). -> Check Gelman's data/NSRI data.

- Introduce case study and set scope of this tutorial: We're providing an overview of implementations. It's up-to the reader to decide which strategy suits their data. So we won't go into detail for the different methods (and equations). This paper is just a software tutorial. We'll keep it practical. -> ADD: some kind of help function that suggests a suitable pred matrix to the user, given a certain analysis model.

@@

## Case study

- We'll use the IMPACT data (`metamisc::impact`) and a MAR/MNAR version of the `mice::popmis` data (i.e., a variation on the Hox (2010) popularity data, where the missingness in the variables is either missing at random (MAR) or missing not at random (MNAR)).

- `impact` is traumatic brain injury data with patients clustered in studies, $n_{\text{participants}} = 11022$ and $n_{\text{clusters}} = 15$, on the following 11 variables:
    * `name` Name of the study,
    * `type` Type of study (RCT: randomized controlled trial, OBS: observational cohort),
    * `age` Age of the patient,
    * `motor_score` Glasgow Coma Scale motor score,
    * `pupil` Pupillary reactivity,
    * `ct` Marshall Computerized Tomography classification,
    * `hypox` Hypoxia (0=no, 1=yes),
    * `hypots` Hypotension (0=no, 1=yes),
    * `tsah` Traumatic subarachnoid hemorrhage (0=no, 1=yes),
    * `edh` Epidural hematoma (0=no, 1=yes),
    * `mort` 6-month mortality (0=alive, 1=dead).

@@

- `popNCR` is a simulated dataset with pupils clustered in classes, $n_{\text{participants}} = 2000$, $n_{\text{clusters}} = 100$, on 7 variables: 
    * `pupil`	Pupil number within class,
    * `class`	Class number,
    * `extrav`	Pupil extraversion,
    * `sex`	Pupil gender,
    * `texp`	Teacher experience (years),
    * `popular`	Pupil popularity,
    * `popteach`	Teacher popularity.

@@

- Pooling 'regular' parameters vs more 'exotic' parameters (SE of residual errors, or autocorreletion)

- ADD: export mids objects to other packages like `lme4` or `coxme`(?)

>>>>

- What is heterogeneity? I.e. variability within studies vs. variability between studies

@@

- What methods are required to analyze multilevel data?	Add references, e.g. @hox17 and @jong21. At least explain difference random effects for intercept term, predictor effects, and/or variance residual error.

@@

-	What kinds of missingness are there? ADD: missingness mechanisms here. See e.g. @yuce08 and @hox15.

@@

- What types of multilevel methods are available? General overview of approaches, see @audi18 and @grun18. E.g., imputation of study level versus patient-level covariates, and one-stage imputation versus two-stage imputation methods.

@@

- We focus on the workflow for conditional modeling (not JOMO) in `mice`. Refer to other packages: `mitml`, `miceadds`, `mdmb`.

- Case study options: `metamisc::impact` (real IPD on traumatic brain injuries, without `NA`s), `mice::popularity` (simulated data on school kids, with MNAR/MAR mixture). TODO: Check example data Gelman.

- Introduce case study and set scope of this tutorial: We're providing an overview of implementations. It's up-to the reader to decide which strategy suits their data. So we won't go into detail for the different methods (and equations). This paper is just a software tutorial. We'll keep it practical. -> ADD: some kind of help function that suggests a suitable predictor matrix to the user, given a certain analysis model.

@@

We'll use the IMPACT data (`metamisc::impact`) and a MAR/MNAR version of the `mice::popmis` data (i.e., a variation on the Hox (2010) popularity data, where the missingness in the variables is either missing at random (MAR) or missing not at random (MNAR)).

## Case study I: IMPACT

`impact` is traumatic brain injury data with patients clustered in studies, $n_{\text{participants}} = 11022$ and $n_{\text{clusters}} = 15$, on the following 11 variables:
    * `name` Name of the study,
    * `type` Type of study (RCT: randomized controlled trial, OBS: observational cohort),
    * `age` Age of the patient,
    * `motor_score` Glasgow Coma Scale motor score,
    * `pupil` Pupillary reactivity,
    * `ct` Marshall Computerized Tomography classification,
    * `hypox` Hypoxia (0=no, 1=yes),
    * `hypots` Hypotension (0=no, 1=yes),
    * `tsah` Traumatic subarachnoid hemorrhage (0=no, 1=yes),
    * `edh` Epidural hematoma (0=no, 1=yes),
    * `mort` 6-month mortality (0=alive, 1=dead).

@@

## Case Study II: Popularity

`popNCR` is a simulated dataset with pupils clustered in classes, $n_{\text{participants}} = 2000$, $n_{\text{clusters}} = 100$, on 7 variables: 
    * `pupil`	Pupil number within class,
    * `class`	Class number,
    * `extrav`	Pupil extraversion,
    * `sex`	Pupil gender,
    * `texp`	Teacher experience (years),
    * `popular`	Pupil popularity,
    * `popteach`	Teacher popularity.

@@

- Pooling 'regular' parameters vs more 'exotic' parameters (SE of residual errors, or autocorrelation)

- ADD: export `mids` objects to other packages like `lme4` or `coxme`?

### 2021-11-04 Manuscript/Manuscript.tex
Add some more references to manuscript

\begin{itemize}
\item
  What is clustering/multilevel data? In this paper, we discuss grouped
  observations, not longitudinal data (within-patient clustering).
  -\textgreater{} ADD: timeseries also in Discussion section.
\item
  What do we mean by clustering? In the medical field: Clustering by
  studies (IPDMA), hospitals in registries, multi-center studies etc. In
  other fields: e.g.~official stats clustering at country-level, or
  social sciences clustering at school-level (related to the sampling
  design).
\item
  What is heterogeniety? I.e. variability within studies vs.~variability
  between studies
\item
  What does multilevel data look like? ADD: figure to show difference
  between patient-level datapoints vs cluster-level datapoints. Maybe
  also add different data frame formats (or just explain in text that
  there's long and wide formats).
\item
  What methods are required to analyze multilevel data? Add references,
  e.g. \citet{hox17}. At least explain difference random effects for
  intercept term, predictor effects, and/or variance residual error.
\end{itemize}

@@

\begin{itemize}
\item
  What kinds of missingness are there? ADD: missingness mechanisms here.
\item
  Why are standard (ad hoc) missing data methods not well suited?
\item
  What types of multilevel methods are available? General overview of
  approaches, see \citet{audi18} \citet{grun18}. E.g., imputation of
  study level versus patient-level covariates, and one-stage imputation
  versus two-stage imputation methods.
\item
  Additional difficulty that is addressed in this tutorial: MNAR data.
\end{itemize}

@@

\begin{itemize}
\item
  Provide practical guidelines with code snippets for imputation of
  incomplete multilevel data.
\item
  We focus on the workflow for conditional modeling (not JOMO) in
  \texttt{mice}. Refer to other packages: \texttt{mitml},
  \texttt{miceadds}.
\item
  Case study options: \texttt{metamisc::impact} (real data on traumatic
  brain injuries, IPD), \texttt{mice::popularity} (simulated data with
  MNAR/MAR mixture, schools). -\textgreater{} Check Gelman's data/NSRI
  data.
\item
  Introduce case study and set scope of this tutorial: We're providing
  an overview of implementations. It's up-to the reader to decide which
  strategy suits their data. So we won't go into detail for the
  different methods (and equations). This paper is just a software
  tutorial. We'll keep it practical. -\textgreater{} ADD: some kind of
  help function that suggests a suitable pred matrix to the user, given
  a certain analysis model.
\end{itemize}

@@

\hypertarget{case-study}{%
\subsection{Case study}\label{case-study}}

\begin{itemize}
\item
  We'll use the IMPACT data (\texttt{metamisc::impact}) and a MAR/MNAR
  version of the \texttt{mice::popmis} data (i.e., a variation on the
  Hox (2010) popularity data, where the missingness in the variables is
  either missing at random (MAR) or missing not at random (MNAR)).
\item
  \texttt{impact} is traumatic brain injury data with patients clustered
  in studies, \(n_{\text{participants}} = 11022\) and
  \(n_{\text{clusters}} = 15\), on the following 11 variables:

  \begin{itemize}
  \tightlist
  \item
    \texttt{name} Name of the study,
  \item
    \texttt{type} Type of study (RCT: randomized controlled trial, OBS:
    observational cohort),
  \item
    \texttt{age} Age of the patient,
  \item
    \texttt{motor\_score} Glasgow Coma Scale motor score,
  \item
    \texttt{pupil} Pupillary reactivity,
  \item
    \texttt{ct} Marshall Computerized Tomography classification,
  \item
    \texttt{hypox} Hypoxia (0=no, 1=yes),
  \item
    \texttt{hypots} Hypotension (0=no, 1=yes),
  \item
    \texttt{tsah} Traumatic subarachnoid hemorrhage (0=no, 1=yes),
  \item
    \texttt{edh} Epidural hematoma (0=no, 1=yes),
  \item
    \texttt{mort} 6-month mortality (0=alive, 1=dead).
  \end{itemize}
\end{itemize}

\begin{CodeChunk}
\begin{CodeInput}
R> # load data
R> data("impact")
R> # # descriptive statistics
R> # by(impact, impact$name, summary) 
R> # psych::describe(impact)[,c(2:5,8:9)]
R> # missingness
R> md_pat(impact)
\end{CodeInput}
\begin{CodeOutput}
 /\     /\
{  `---'  }
{  O   O  }
==>  V <==  No need for mice. This data set is completely observed.
 \  \|/  /
  `-----'
\end{CodeOutput}
\begin{CodeOutput}
      name type age motor_score pupil ct hypox hypots tsah edh mort  
11022    1    1   1           1     1  1     1      1    1   1    1 0
         0    0   0           0     0  0     0      0    0   0    0 0
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{itemize}
\tightlist
\item
  \texttt{popNCR} is a simulated dataset with pupils clustered in
  classes, \(n_{\text{participants}} = 2000\),
  \(n_{\text{clusters}} = 100\), on 7 variables:

  \begin{itemize}
  \tightlist
  \item
    \texttt{pupil} Pupil number within class,
  \item
    \texttt{class} Class number,
  \item
    \texttt{extrav} Pupil extraversion,
  \item
    \texttt{sex} Pupil gender,
  \item
    \texttt{texp} Teacher experience (years),
  \item
    \texttt{popular} Pupil popularity,
  \item
    \texttt{popteach} Teacher popularity.
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\item
  Analysis of scientific interest.
\item
  Pooling using \texttt{mitml}.
\item
  Pooling `regular' parameters vs more `exotic' parameters (SE of
  residual errors, or autocorreletion)
\item
  ADD: export mids objects to other packages like \texttt{lme4} or
  \texttt{coxme}(?)
\end{itemize}

>>>>

\begin{itemize}
\item
  What is clustering/multilevel data? In this paper, we discuss grouped
  observations, not longitudinal data (within-patient clustering).
  -\textgreater{} ADD: timeseries also in Discussion section.
\item
  What do we mean by clustering? In the medical field: Clustering by
  studies (IPDMA), hospitals in registries, multi-center studies etc. In
  other fields: e.g.~official stats clustering at country-level, or
  social sciences clustering at school-level (related to the sampling
  design).
\item
  What is heterogeneity? I.e. variability within studies vs.~variability
  between studies
\item
  What does multilevel data look like? ADD: figure to show difference
  between patient-level datapoints vs cluster-level datapoints. Maybe
  also add different data frame formats (or just explain in text that
  there's long and wide formats).
\item
  What methods are required to analyze multilevel data? Add references,
  e.g. \citet{hox17} and \citet{jong21}. At least explain difference
  random effects for intercept term, predictor effects, and/or variance
  residual error.
\end{itemize}

@@

\begin{itemize}
\item
  What kinds of missingness are there? ADD: missingness mechanisms here.
  See e.g. \citet{yuce08} and \citet{hox15}.
\item
  Why are standard (ad hoc) missing data methods not well suited?
\item
  What types of multilevel methods are available? General overview of
  approaches, see \citet{audi18} and \citet{grun18}. E.g., imputation of
  study level versus patient-level covariates, and one-stage imputation
  versus two-stage imputation methods.
\item
  Additional difficulty that is addressed in this tutorial: MNAR data.
\end{itemize}

@@

\begin{itemize}
\item
  Provide practical guidelines with code snippets for imputation of
  incomplete multilevel data.
\item
  We focus on the workflow for conditional modeling (not JOMO) in
  \texttt{mice}. Refer to other packages: \texttt{mitml},
  \texttt{miceadds}, \texttt{mdmb}.
\item
  Case study options: \texttt{metamisc::impact} (real IPD on traumatic
  brain injuries, without \texttt{NA}s), \texttt{mice::popularity}
  (simulated data on school kids, with MNAR/MAR mixture). TODO: Check
  example data Gelman.
\item
  Introduce case study and set scope of this tutorial: We're providing
  an overview of implementations. It's up-to the reader to decide which
  strategy suits their data. So we won't go into detail for the
  different methods (and equations). This paper is just a software
  tutorial. We'll keep it practical. -\textgreater{} ADD: some kind of
  help function that suggests a suitable predictor matrix to the user,
  given a certain analysis model.
\end{itemize}

@@

We'll use the IMPACT data (\texttt{metamisc::impact}) and a MAR/MNAR
version of the \texttt{mice::popmis} data (i.e., a variation on the Hox
(2010) popularity data, where the missingness in the variables is either
missing at random (MAR) or missing not at random (MNAR)).

\hypertarget{case-study-i-impact}{%
\subsection{Case study I: IMPACT}\label{case-study-i-impact}}

\texttt{impact} is traumatic brain injury data with patients clustered
in studies, \(n_{\text{participants}} = 11022\) and
\(n_{\text{clusters}} = 15\), on the following 11 variables: *
\texttt{name} Name of the study, * \texttt{type} Type of study (RCT:
randomized controlled trial, OBS: observational cohort), * \texttt{age}
Age of the patient, * \texttt{motor\_score} Glasgow Coma Scale motor
score, * \texttt{pupil} Pupillary reactivity, * \texttt{ct} Marshall
Computerized Tomography classification, * \texttt{hypox} Hypoxia (0=no,
1=yes), * \texttt{hypots} Hypotension (0=no, 1=yes), * \texttt{tsah}
Traumatic subarachnoid hemorrhage (0=no, 1=yes), * \texttt{edh} Epidural
hematoma (0=no, 1=yes), * \texttt{mort} 6-month mortality (0=alive,
1=dead).

\begin{CodeChunk}
\begin{CodeInput}
R> # load data
R> data("impact")
R> # # descriptive statistics
R> # by(impact, impact$name, summary) 
R> # psych::describe(impact)[,c(2:5,8:9)]
R> # missingness
R> md_pat(impact)
\end{CodeInput}
\begin{CodeOutput}
 /\     /\
{  `---'  }
{  O   O  }
==>  V <==  No need for mice. This data set is completely observed.
 \  \|/  /
  `-----'
\end{CodeOutput}

\begin{center}\includegraphics{Manuscript_files/figure-latex/impact-1} \end{center}

@@

\hypertarget{case-study-ii-popularity}{%
\subsection{Case Study II: Popularity}\label{case-study-ii-popularity}}

\texttt{popNCR} is a simulated dataset with pupils clustered in classes,
\(n_{\text{participants}} = 2000\), \(n_{\text{clusters}} = 100\), on 7
variables: * \texttt{pupil} Pupil number within class, * \texttt{class}
Class number, * \texttt{extrav} Pupil extraversion, * \texttt{sex} Pupil
gender, * \texttt{texp} Teacher experience (years), * \texttt{popular}
Pupil popularity, * \texttt{popteach} Teacher popularity.

@@

\begin{itemize}
\item
  Analysis of scientific interest.
\item
  Pooling using \texttt{mitml}.
\item
  Pooling `regular' parameters vs more `exotic' parameters (SE of
  residual errors, or autocorrelation)
\item
  ADD: export \texttt{mids} objects to other packages like \texttt{lme4}
  or \texttt{coxme}?
\end{itemize}
